FOODiE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2015-10-07 papers/announcement/manuscript.tex
minor improvements into manuscript

\subsubsection{Errors Analysis of the FOODiE Oscillation test results}

@@

    \cmidrule{2-6}
    \multirow{6}{*}{leapfrog 2 steps}
                 &   5000.0        &  0.151E+02    &  0.152E+02    & /             & /             \\
                 &   2500.0        &  0.500E+01    &  0.504E+01    & 1.60          & 1.59          \\
                 &   1250.0        &  0.190E+01    &  0.193E+01    & 1.40          & 1.39          \\
                 &    625.0        &  0.181E+01    &  0.181E+01    & 0.07          & 0.09          \\
                 &    320.0        &  0.157E+01    &  0.157E+01    & 0.21          & 0.21          \\
                 &    100.0        &  0.100E+01    &  0.100E+01    & 0.39          & 0.39          \\
    \cmidrule{2-6}
    \multirow{6}{*}{low storage Runge-Kutta 5 stages}
                 &   5000.0        &  0.120E+00    &  0.122E+00    & /             & /             \\
                 &   2500.0        &  0.106E-01    &  0.107E-01    & 3.51          & 3.51          \\
                 &   1250.0        &  0.935E-03    &  0.947E-03    & 3.50          & 3.50          \\
                 &    625.0        &  0.826E-04    &  0.836E-04    & 3.50          & 3.50          \\
                 &    320.0        &  0.793E-05    &  0.803E-05    & 3.50          & 3.50          \\
                 &    100.0        &  0.135E-06    &  0.137E-06    & 3.50          & 3.50          \\
    \cmidrule{2-6}
    \multirow{6}{*}{TVD/SSP Runge-Kutta 2 stages}
                 &   5000.0        &  0.316E+02    &  0.319E+02    & /             & /             \\
                 &   2500.0        &  0.892E+01    &  0.894E+01    & 1.83          & 1.84          \\
                 &   1250.0        &  0.301E+01    &  0.305E+01    & 1.57          & 1.55          \\
                 &    625.0        &  0.106E+01    &  0.107E+01    & 1.51          & 1.51          \\
                 &    320.0        &  0.387E+00    &  0.392E+00    & 1.50          & 1.50          \\
                 &    100.0        &  0.676E-01    &  0.685E-01    & 1.50          & 1.50          \\
    \cmidrule{2-6}
    \multirow{6}{*}{TVD/SSP Runge-Kutta 3 stages}
                 &   5000.0        &  0.255E+01    &  0.252E+01    & /             & /             \\
                 &   2500.0        &  0.523E+00    &  0.516E+00    & 2.28          & 2.29          \\
                 &   1250.0        &  0.944E-01    &  0.931E-01    & 2.47          & 2.47          \\
                 &    625.0        &  0.167E-01    &  0.165E-01    & 2.50          & 2.50          \\
                 &    320.0        &  0.314E-02    &  0.310E-02    & 2.50          & 2.50          \\
                 &    100.0        &  0.171E-03    &  0.169E-03    & 2.50          & 2.50          \\
    \cmidrule{2-6}
    \multirow{6}{*}{TVD/SSP Runge-Kutta 5 stages}
                 &   5000.0        &  0.139E+00    &  0.141E+00    & /             & /             \\
                 &   2500.0        &  0.122E-01    &  0.124E-01    & 3.50          & 3.50          \\
                 &   1250.0        &  0.108E-02    &  0.110E-02    & 3.50          & 3.50          \\
                 &    625.0        &  0.956E-04    &  0.969E-04    & 3.50          & 3.50          \\
                 &    320.0        &  0.937E-05    &  0.949E-05    & 3.47          & 3.47          \\
                 &    100.0        &  0.512E-06    &  0.519E-06    & 2.50          & 2.50          \\
    \bottomrule
  \end{tabular}}
\end{table}

@@

Note that in the summary.log file into error analysis directory I have computed also the product of $f*\Delta t$ in order to verify when the instability arises: for AB-2 and AB-4 the solution with $\Delta t=5000 \rightarrow f*\Delta t=0.5$ results to be unstable while for AB-3 it is stable, but the amplitude is dumped resulting into large errors. This is consistent with Duran results, where AB-3 is said to be unstable for $ f*\Delta t>0.724$.
}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.70\textwidth]{errors-analysis/oscillation/errors_analysis-oscillation-leapfrog.png}
  \caption{Oscillation equations solutions computed by means of leapfrog 2 steps solver}\label{fig:results-oscillation-leapfrog}
\end{figure}

>>>>

\paragraph{FOODiE numerical integration}

Using the above discussed \emph{oscillation} type it is very easy to solve IVP \ref{eq:oscillation} by means of FOODiE library. Listing \ref{list:oscillation_leapfrog} presents the numerical integration of system \ref{eq:oscillation} by means of the Leapfrog RAW-filtered method. In the example, the integration is performed with $10^4$ steps with a fixed $\Delta t=10^2$ until the time $t=10^6$ is reached. The example shows also that for starting a multi-step scheme such as the Leapfrog one a lower-oder or equivalent order one-scheme is necessary: in the example the first 2 steps are computed by means of one-step TVD/SSP Runge-Kutta 2-stages schemes. Note that the memory registers for storing the Runge-Kutta stages and the RAW filter displacement must be handled by the client code. Listing \ref{list:oscillation_leapfrog} demonstrates how it is simple, clear and concise to solve a IVP by FOODiE solvers. Moreover, it proves how it is simple and effective to apply different solvers in a coupled algorithm, that greatly simplify the development of new hybrid solvers for self-adaptive time step size.

\begin{lstlisting}[firstnumber=1,style=code,caption={numerical integration of the \emph{oscillation} system by means of Leapfrog RAW-filtered method},label={list:oscillation_leapfrog}]
use foodie, only: leapfrog_integrator, tvd_runge_kutta_integrator
type(leapfrog_integrator)        :: lf_integrator ! Leapfrog integrator.
type(tvd_runge_kutta_integrator) :: rk_integrator ! Runge-Kutta integrator.
type(oscillation)                :: rk_stage(1:2) ! Runge-Kutta stages.
type(oscillation)                :: oscillator    ! Oscillation field.
type(oscillation)                :: filter        ! Filter displacement.
integer                          :: step          ! Time steps counter.
real                             :: Dt            ! Time step.
call lf_integrator%init()
call rk_integrator%init(stages=2)
call oscillator%init(initial_state=[0.0,1.0], f=10e4, steps=2)
Dt = 100.0
do step=1, 10000
  if (2>=step) then
    call rk_integrator%integrate(U=oscillator, stage=rk_stage, Dt=Dt, t=step*Dt)
  else
    call lf_integrator%integrate(U=oscillator, filter=filter, Dt=Dt, t=step*Dt)
  endif
enddo
call print_results(U=oscillator)
\end{lstlisting}

\paragraph{FOODiE integration}
\subsubsection{Errors Analysis of the FOODiE Oscillation test results}

@@

    \cmidrule{2-6}
    \multirow{6}{*}{Leapfrog RAW-filtered}
                 &   5000.0        &  0.151E+02    &  0.152E+02    & /             & /             \\
                 &   2500.0        &  0.500E+01    &  0.504E+01    & 1.60          & 1.59          \\
                 &   1250.0        &  0.190E+01    &  0.193E+01    & 1.40          & 1.39          \\
                 &    625.0        &  0.181E+01    &  0.181E+01    & 0.07          & 0.09          \\
                 &    320.0        &  0.157E+01    &  0.157E+01    & 0.21          & 0.21          \\
                 &    100.0        &  0.100E+01    &  0.100E+01    & 0.39          & 0.39          \\
    \cmidrule{2-6}
    \multirow{6}{*}{low storage Runge-Kutta 5 stages}
                 &   5000.0        &  0.120E+00    &  0.122E+00    & /             & /             \\
                 &   2500.0        &  0.106E-01    &  0.107E-01    & 3.51          & 3.51          \\
                 &   1250.0        &  0.935E-03    &  0.947E-03    & 3.50          & 3.50          \\
                 &    625.0        &  0.826E-04    &  0.836E-04    & 3.50          & 3.50          \\
                 &    320.0        &  0.793E-05    &  0.803E-05    & 3.50          & 3.50          \\
                 &    100.0        &  0.135E-06    &  0.137E-06    & 3.50          & 3.50          \\
    \cmidrule{2-6}
    \multirow{6}{*}{TVD/SSP Runge-Kutta 2 stages}
                 &   5000.0        &  0.316E+02    &  0.319E+02    & /             & /             \\
                 &   2500.0        &  0.892E+01    &  0.894E+01    & 1.83          & 1.84          \\
                 &   1250.0        &  0.301E+01    &  0.305E+01    & 1.57          & 1.55          \\
                 &    625.0        &  0.106E+01    &  0.107E+01    & 1.51          & 1.51          \\
                 &    320.0        &  0.387E+00    &  0.392E+00    & 1.50          & 1.50          \\
                 &    100.0        &  0.676E-01    &  0.685E-01    & 1.50          & 1.50          \\
    \cmidrule{2-6}
    \multirow{6}{*}{TVD/SSP Runge-Kutta 3 stages}
                 &   5000.0        &  0.255E+01    &  0.252E+01    & /             & /             \\
                 &   2500.0        &  0.523E+00    &  0.516E+00    & 2.28          & 2.29          \\
                 &   1250.0        &  0.944E-01    &  0.931E-01    & 2.47          & 2.47          \\
                 &    625.0        &  0.167E-01    &  0.165E-01    & 2.50          & 2.50          \\
                 &    320.0        &  0.314E-02    &  0.310E-02    & 2.50          & 2.50          \\
                 &    100.0        &  0.171E-03    &  0.169E-03    & 2.50          & 2.50          \\
    \cmidrule{2-6}
    \multirow{6}{*}{TVD/SSP Runge-Kutta 5 stages}
                 &   5000.0        &  0.139E+00    &  0.141E+00    & /             & /             \\
                 &   2500.0        &  0.122E-01    &  0.124E-01    & 3.50          & 3.50          \\
                 &   1250.0        &  0.108E-02    &  0.110E-02    & 3.50          & 3.50          \\
                 &    625.0        &  0.956E-04    &  0.969E-04    & 3.50          & 3.50          \\
                 &    320.0        &  0.937E-05    &  0.949E-05    & 3.47          & 3.47          \\
                 &    100.0        &  0.512E-06    &  0.519E-06    & 2.50          & 2.50          \\
    \bottomrule
  \end{tabular}}
\end{table}

@@

Note that in the summary.log file into error analysis directory I have computed also the product of $f*\Delta t$ in order to verify when the instability arises.

For AB-2 and AB-4 the solution with $\Delta t=5000 \rightarrow f*\Delta t=0.5$ results to be unstable (the computational modes are not dumped) while for AB-3 it is stable, but the (physical mode) amplitude is dumped resulting into large errors. This is consistent with Duran results, where AB-3 is said to be unstable for $ f*\Delta t>0.724$.

The AB-2 and AB-4 solutions show a phase error that decreases as the time resolution increases. The AB-3 has no phase error and the computational modes are dumped for all $\Delta t$ exercised.
}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.70\textwidth]{errors-analysis/oscillation/errors_analysis-oscillation-leapfrog.png}
  \caption{Oscillation equations solutions computed by means of Leapfrog RAW-filtered solver}\label{fig:results-oscillation-leapfrog}
\end{figure}
